---
title: "ResearchPilot Phase 1 — Exploratory Data Analysis"
subtitle: "final_dataset.csv"
author: "ResearchPilot"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

# Setup

```{r libraries}
library(tidyverse)
library(skimr)
library(janitor)
library(scales)
library(GGally)

theme_set(theme_minimal(base_size = 12))
```

```{r paths}
find_project_root <- function() {
  path <- normalizePath(getwd(), winslash = "/", mustWork = TRUE)
  for (i in seq_len(6)) {
    candidate <- file.path(path, "data", "final", "final_dataset.csv")
    if (file.exists(candidate)) return(path)
    parent <- dirname(path)
    if (identical(parent, path)) break
    path <- parent
  }
  stop("Project root with data/final/final_dataset.csv not found.")
}

project_root <- find_project_root()
input_csv <- file.path(project_root, "data", "final", "final_dataset.csv")
viz_dir <- file.path(project_root, "r", "visualizations")
dir.create(viz_dir, recursive = TRUE, showWarnings = FALSE)
```

# Load data

```{r load}
papers <- read_csv(input_csv, show_col_types = FALSE) %>%
  mutate(
    publication_year = as.integer(publication_year),
    cited_by_count = as.numeric(cited_by_count),
    is_open_access = as.logical(is_open_access),
    has_fulltext = as.logical(has_fulltext),
    recent_paper = as.integer(recent_paper),
    has_doi = as.integer(has_doi),
    oa_status = factor(oa_status),
    oa_category = factor(
      oa_category,
      levels = c("fully_open", "partially_open", "closed", "unknown")
    )
  )

glimpse(papers)
```

# Dataset overview

```{r overview}
dim(papers)
skim(papers)
```

# Missing values

```{r missing}
missing_tbl <- papers %>%
  summarise(across(everything(), ~ sum(is.na(.x)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "n_missing") %>%
  mutate(pct_missing = round(100 * n_missing / nrow(papers), 2)) %>%
  arrange(desc(n_missing))

missing_tbl
```

# Papers by year and age

```{r year-age}
papers %>% count(publication_year)

ggplot(papers, aes(x = factor(publication_year))) +
  geom_bar(fill = "#1B9E77") +
  labs(title = "Papers by Publication Year", x = "Year", y = "Count")

ggplot(papers, aes(x = paper_age)) +
  geom_bar(fill = "#7570B3") +
  labs(title = "Paper Age Distribution", x = "Age (years)", y = "Count")
```

# Citation analysis

```{r citations}
papers %>%
  summarise(
    mean = mean(cited_by_count),
    median = median(cited_by_count),
    max = max(cited_by_count),
    mean_per_year = mean(citation_per_year),
    median_log = median(citation_log)
  )

ggplot(papers, aes(x = cited_by_count)) +
  geom_histogram(bins = 40, fill = "#D95F02", color = "white") +
  scale_x_continuous(labels = comma) +
  labs(title = "Raw Citation Counts", x = "cited_by_count", y = "Count")

ggplot(papers, aes(x = citation_log)) +
  geom_histogram(bins = 40, fill = "#E7298A", color = "white") +
  labs(title = "Log Citations", x = "citation_log", y = "Count")

ggplot(papers, aes(x = factor(publication_year), y = citation_log)) +
  geom_boxplot(fill = "#A6761D", alpha = 0.8) +
  labs(title = "Log Citations by Year", x = "Year", y = "citation_log")
```

# Text length and topical richness

```{r text-topics}
ggplot(papers, aes(x = title_length)) +
  geom_histogram(bins = 35, fill = "#1F78B4", color = "white") +
  labs(title = "Title Length", x = "Characters", y = "Count")

ggplot(papers, aes(x = abstract_length)) +
  geom_histogram(bins = 35, fill = "#33A02C", color = "white") +
  labs(title = "Abstract Length", x = "Characters", y = "Count")

ggplot(papers, aes(x = keyword_count, y = concept_count)) +
  geom_point(alpha = 0.25, color = "#7570B3") +
  geom_smooth(method = "lm", se = TRUE, color = "#E7298A") +
  labs(title = "Keywords vs Concepts")
```

# Open access

```{r oa}
papers %>%
  count(oa_category) %>%
  mutate(pct = round(100 * n / sum(n), 1))

ggplot(papers, aes(x = oa_category, fill = oa_category)) +
  geom_bar(show.legend = FALSE) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "OA Category", x = NULL, y = "Count")

ggplot(papers, aes(x = oa_category, y = citation_log, fill = oa_category)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Log Citations by OA Category")
```

# Correlation heatmap

```{r correlation, fig.width=9, fig.height=7}
numeric_feats <- papers %>%
  select(
    publication_year, cited_by_count, paper_age, title_length,
    abstract_length, keyword_count, concept_count, citation_per_year,
    citation_log, recent_paper, has_doi
  )

corr_long <- as.data.frame(as.table(cor(numeric_feats, use = "pairwise.complete.obs"))) %>%
  rename(feature_x = Var1, feature_y = Var2, correlation = Freq)

ggplot(corr_long, aes(feature_x, feature_y, fill = correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", correlation)), size = 2.5) +
  scale_fill_gradient2(low = "#B2182B", mid = "white", high = "#2166AC", midpoint = 0) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Numeric Feature Correlations", x = NULL, y = NULL)
```

# Top cited papers

```{r top-cited}
papers %>%
  arrange(desc(cited_by_count)) %>%
  select(title, publication_year, cited_by_count, citation_per_year, oa_category) %>%
  slice_head(n = 10) %>%
  knitr::kable()
```

# Interpretation prompts

1. How are papers distributed across years?
2. How skewed are citations, and does `citation_log` help?
3. Do OA categories differ in impact?
4. Which engineered features look useful for Phase 2 ML?
5. What limitations or biases should be documented?
